---
title: "Reproducing the figures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(pgsbridge)
```

This vignette shows what the package computes, then how the manuscript's
figures are produced from it. There are two routes: redrawing the figures from
the pre-computed results, which takes seconds, and regenerating those results
from genotypes, which takes a cluster and several days.

## The ceiling

The best accuracy a polygenic score can attain in a population is the share of
that population's trait variance that is genetic. Three quantities set it: the
additive genetic variance fixed by the allele-frequency spectrum, the average
self-kinship summarising within-population structure, and the non-genetic
variance.

```{r}
# equal genetic and non-genetic variance, no within-population structure
prediction_accuracy(sigma2 = 1, phi_bar = 0.5, tau2 = 1)
```

Only the product of the additive genetic variance and the average self-kinship
matters, never either alone. Doubling one and halving the other leaves the
ceiling unchanged:

```{r}
prediction_accuracy(sigma2 = 2.0, phi_bar = 0.25, tau2 = 1)
prediction_accuracy(sigma2 = 0.5, phi_bar = 1.00, tau2 = 1)
```

The non-genetic variance acts on its own, which is why environmental variation
alone can move transferability away from one.

## Transferability

Transferability is the ratio of two such ceilings. It is exactly reciprocal:
carrying a score from `A` to `B` and back must give one.

```{r}
forward <- transferability(
    sigma2_A = 1.0, phi_bar_A = 0.5, tau2_A = 1,
    sigma2_B = 1.4, phi_bar_B = 0.5, tau2_B = 1
)
backward <- transferability(
    sigma2_A = 1.4, phi_bar_A = 0.5, tau2_A = 1,
    sigma2_B = 1.0, phi_bar_B = 0.5, tau2_B = 1
)
c(forward = forward, backward = backward, product = forward * backward)
```

That identity is worth remembering when reading simulation output. A single
empirical estimate satisfies it by construction, being a ratio of the same two
accuracies. A *mean over replicates* need not, and when it does not, that is
Jensen's inequality telling you the ratio-of-means estimator is biased, not that
the populations are unusual.

## Measuring it rather than evaluating it

The same quantity can be measured from simulated traits. Choose one effect size
shared by every causal variant so that the trait reaches a target heritability,
simulate, and score the optimal predictor.

```{r}
set.seed(1)

m_loci <- 1000
n_ind <- 5000
p <- runif(m_loci, 0.05, 0.95)
X <- matrix(rbinom(n_ind * m_loci, 2, rep(p, each = n_ind)), n_ind, m_loci)

beta <- rep(shared_effect_size(h2 = 0.4, tau2 = 1, p = p), m_loci)
trait <- simulate_trait(X, beta, alpha = 0, tau2 = 1)
```

The measured accuracy and the closed form evaluated at the realised allele
frequencies agree:

```{r}
c(
    measured = empirical_accuracy(trait$y, trait$g),
    closed_form = prediction_accuracy(
        sigma2 = additive_genetic_variance(beta, allele_frequencies(X)),
        phi_bar = 0.5, tau2 = 1
    )
)
```

## Redrawing the manuscript figures

The repository ships the replicate-level output of every simulation under
`data/`, so the figures need no genotypes and no cluster. From the repository
root:

```
Rscript scripts/figures/figure_01.R
Rscript scripts/figures/figure_02.R
...
Rscript scripts/figures/figure_S8.R
```

Each writes a PDF and a PNG into `output/`. The scripts read their inputs
through `read_transferability_by_heritability()` and `read_rare_variant_sweep()`,
which stack the replicate files and summarise them with `group_summary()`.

`group_summary()` trims replicates outside 1.5 times the interquartile range
within each group before averaging. Transferability is a ratio, and when the
denominator accuracy is small in a replicate the ratio explodes, so a few
replicates would otherwise dominate the mean. The number of replicates kept is
returned as `n`.

## Regenerating the results

`scripts/coalescent/` produces the genotypes: an out-of-Africa coalescent
simulation, a split of the variants into common and rare pools, and a check on
the resulting between-population differentiation.  Its own `README.md` documents
the demographic parameters.  `scripts/simulation/` then simulates traits on
those genotypes, numbered in the order they run.

| Script | Produces |
| --- | --- |
| `coalescent/01_simulate_genotypes.py` | genotypes, written as a VCF |
| `coalescent/02_prepare_genotypes.sh` | PLINK files, split by population |
| `coalescent/03_compute_fst.py` | the pairwise fixation index |
| `simulation/01_theory_validation.R` | Figure 1a |
| `simulation/02_traits_common_variants.R` | Figure 2, S1 Fig, S2 Fig |
| `simulation/03_traits_normal_effects.R` | S2 Fig |
| `simulation/04_traits_rare_and_common.R` | Figure 3, S4 to S6 Figs |
| `simulation/05_heritability_decomposition.R` | Figure 4 |
| `simulation/06_allele_frequencies.R` | S3, S7 Figs |
| `simulation/07_ascertainment_schemes.R` | S8 Fig |

Step 1 of the trait simulations is self-contained and simulates its own
genotypes with `bnpsd`.  Steps 2 to 7 read the PLINK files from step 2 through `load_variant_pools()`, which
returns column indices rather than genotypes: the rare pool runs to millions of
variants and holding it for three populations at once would need tens of
gigabytes. `draw_causal_genotypes()` then reads only the drawn columns.

Step 7 additionally requires that the three per-population panels share one
coded allele, because it adds allele counts across populations to build the
pooled and all-three variant pools. `coalescent/02_prepare_genotypes.sh`
guarantees this with `--keep-allele-order`; step 7 checks it and stops if the
panels disagree.

```{r eval = FALSE}
pools <- load_variant_pools("genotypes")
causal <- draw_causal_genotypes(pools, "rare", 100)
```
